Deciphering the Trends: A Comprehensive Analysis of US Birth Rates and Educational Attainment from 1980 to 2020

### 2024-02-12 outputs/paper/paper.qmd
edit
Improved figures and labels, change to introduction paragraphs

---
title: "Deciphering the Trends: A Comprehensive Analysis of US Birth Rates and Educational Attainment from 1980 to 2020"
author: "Jagpreet Singh, Alexander Guarasci & Jacob Gilbert"
date: "`r format(Sys.Date(), '%B %d, %Y')`" # This will add the current date
date-format: long
bibliography: references.bib
execute:
  echo: false
format: pdf
fig-pos: "H"
abstract: |
 This paper examines the trends in birth rates across the United States from 1980 to 2020, analyzing how factors like marital status, education level, and age group influence fertility decisions.
 Through the examination of comprehensive data, provided by American Economics Association [@citewebsite], we uncovered a significant decline in birth rates post-2007, with marked variations across different demographic groups, indicating a shift towards later-life parenthood and a nuanced relationship between education and fertility. These findings highlight the complex interplay between socioeconomic factors and reproductive choices, shedding light on how societal norms and individual priorities are reshaping family formation patterns.
---

```{r}
knitr::opts_chunk$set(echo = FALSE, results = "hide", message = FALSE, warning = FALSE)
```
Code and data underpinning this paper are available at: https://github.com/AlexanderG123/usa_birthrate. Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: DOI : https://doi.org/10.48152/ssrp-g8j2-3h68

# Introduction

@@

The dataset was obtained from the data section for “The Puzzle of Falling US Birth Rates since the Great Recession” [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite]. Analysis of the data was done using the R programming language [@citeR] in conjunction with the following packages: tidyverse [@tidyverse] and Knitr [@knitr]. The figures were created using ggplot2 [@citeggplot]. We began by constructing a comprehensive line graph illustrating the overarching trend of the birth rate per 1000 women across all women aged 15-44 in the United States from 1980-2020. We then plotted three-line graphs to analyze the patterns in birth rate with relation to three distinct demographic groups defined by marital status, educational level, and age group of the mother. This breakdown facilitated a more nuanced understanding of how birth rates varied within these demographic cohorts for the specified period. This helped us study the overall trend concerning these three demographic groups.

This paper is organized as follows: first, we describe the methodological approach, including data sourcing and analytical tools; then, we present our results in detail; we discuss our findings in the context of previous research, pointing out the limitations of our study as well as its societal implications; finally, we suggest future directions for research, stressing the significance of looking into the underlying causes of changing fertility patterns and their effects on society.

@@

# 2 Data

# 2.1 Description of datasets:

@@

# 2.2 Data cleaning and construction:

@@

# 2.3 Variable explanation:

@@

# 2.4 Measurement:

@@

# 3 Results:

@@

```{r, fig.cap= "Birth rate for women aged 15-44 regressing to a declining linear model"}
#| label: fig-1
library(tidyverse)
data <- read.csv(here::here("inputs/data/fig_1.csv"))

@@

```{r, fig.cap= "Birth rate of different age groups"}
#| label: fig-2
data <- read.csv(here::here("inputs/data/fig_4.csv"))
ggplot() + 
    geom_line(data = data, aes(x = year, y = brate_1519, color = "15-19")) +
    geom_line(data = data, aes(x = year, y = brate_2024, color = "20-24")) +
    geom_line(data = data, aes(x = year, y = brate_2529, color = "25-29")) +
    geom_line(data = data, aes(x = year, y = brate_3034, color = "30-34")) +
    geom_line(data = data, aes(x = year, y = brate_3539, color = "35-39")) +
    geom_line(data = data, aes(x = year, y = brate_4044, color = "40-44")) +
    theme_minimal() +
    labs(title = "US Birth Rates by Age Group (1980-2020)",
         x = "Year",
         y = "Birth Rate",
         color = "Age Group")

@@

```{r, fig.cap= "Birth rate by different education level"}
#| label: fig-3
data <- read.csv(here::here("inputs/data/fig_5.csv"))

ggplot() + 
    geom_line(data = data, aes(x = year, y = brate_lesshs, color = "Less than High School")) +
    geom_line(data = data, aes(x = year, y = brate_hsgrad, color = "High School Graduate")) +
    geom_line(data = data, aes(x = year, y = brate_somecoll, color = "Some College")) +
    geom_line(data = data, aes(x = year, y = brate_colgrad, color = "College Graduate")) +
    theme_minimal() +
    labs(title = "US Birth Rates by Education Level (1992-2019)",
         x = "Year",
         y = "Birth Rate",
         color = "Education Level")
```
# 3.3 Marital Status vs Birth Rate

@@

```{r, fig.cap= "Birth rate of married vs unmarried women"}
#| label: fig-4
data <- read.csv(here::here("inputs/data/fig_2e.csv"))

ggplot(data, aes(x = year)) + 
    geom_line(aes(y = brate_married, color = "Married")) +
    geom_line(aes(y = brate_unmarried, color = "Unmarried")) +
    theme_minimal() +
    labs(title = "US Birth Rates by Marital Status (1980-2020)",
         x = "Year",
         y = "Birth Rate",
         color = "Marital Status")

```
# 4 Discussion:

@@

>>>>

---
title: "Deciphering the Trends: A Comprehensive Analysis of US Birth Rates and Educational Attainment from 1980 to 2020"
author: "Jagpreet Singh, Alexander Guarasci & Jacob Gilbert"
thanks: "Code and data underpinning this paper are available at: https://github.com/AlexanderG123/usa_birthrate. Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: DOI : https://doi.org/10.48152/ssrp-g8j2-3h68"
date: "`r format(Sys.Date(), '%B %d, %Y')`" # This will add the current date
date-format: long
bibliography: references.bib
execute:
  echo: false
format: pdf
fig-pos: "H"
abstract: |
 This paper examines the trends in birth rates across the United States from 1980 to 2020, analyzing how factors like marital status, education level, and age group influence fertility decisions. Through the examination of comprehensive data, provided by American Economics Association [@citewebsite], we uncovered a significant decline in birth rates post-2007, with marked variations across different demographic groups, indicating a shift towards later-life parenthood and a nuanced relationship between education and fertility. These findings highlight the complex interplay between socioeconomic factors and reproductive choices, shedding light on how societal norms and individual priorities are reshaping family formation patterns.
---

```{r}
knitr::opts_chunk$set(echo = FALSE, results = "hide", message = FALSE, warning = FALSE)
```

# Introduction {#sec-intro}

@@

Analysis of data “The Puzzle of Falling US Birth Rates since the Great Recession” [@citepaper], a paper from the Journal of Economic Perspectives [@citewebsite], was done using the R programming language [@citeR] in conjunction with the following packages: tidyverse [@tidyverse] and Knitr [@knitr]. The figures were created using ggplot2 [@citeggplot]. Data points on birth rates per 1000 women was used to observe overarching birth patterns in the US from women ages 15-44 from 1980-2020. Demographic variables were also obtained from the data providing functionality for further analysis. Birth rates by age group, marital status and education level were available and selected as independant variables from the data. This breakdown facilitated a more nuanced understanding of how birth rates varied within these demographic cohorts for the specified period.

This paper is organized as follows: first discussing the data @sec-data, we describing the methodological approach, including data sourcing and analytical tools; then, presenting our results in the detail @sec-results; we discuss our findings in the context of previous research, pointing out the limitations of our study as well as its societal implications; finally, In the discussion @sec-discussion is suggestion for future directions of research, stressing the significance of looking into the underlying causes of changing fertility patterns and their effects on society.

@@

# 2 Data {#sec-data}

# 2.1 Description of datasets

@@

# 2.2 Data cleaning and construction

@@

# 2.3 Variable explanation

@@

# 2.4 Measurement

@@

# 3 Results: {#sec-results}

@@

```{r, fig.cap= "Birth rate for women aged 15-44 regressed on a declining linear model"}
#| label: fig-1
library(tidyverse)
data <- read.csv(here::here("inputs/data/fig_1.csv"))

@@

```{r, fig.cap= "Birth rate of different age groups, showing decline in younger age group while an increase in older groups (30+)"}
#| label: fig-2
data <- read.csv(here::here("inputs/data/fig_4.csv"))
ggplot() + 
  geom_line(data = data, aes(x = year, y = brate_1519, color = "15-19")) +
  geom_line(data = data, aes(x = year, y = brate_2024, color = "20-24")) +
  geom_line(data = data, aes(x = year, y = brate_2529, color = "25-29")) +
  geom_line(data = data, aes(x = year, y = brate_3034, color = "30-34")) +
  geom_line(data = data, aes(x = year, y = brate_3539, color = "35-39")) +
  geom_line(data = data, aes(x = year, y = brate_4044, color = "40-44")) +
  theme_minimal() +
  labs(title = "US Birth Rates by Age Group (1980-2020)",
       x = "Year",
       y = "Birth Rate",
       color = "Age Group") +
  theme(legend.position = "bottom",
        legend.title = element_text(size=8),
        legend.key.width = unit(0.5, "cm")) +
  guides(color = guide_legend(nrow = 1))

@@

```{r, fig.cap= "Birth rate by different education level. Most volatility seen in the lowest educated group, and increasing rate observed in highest educated group"}
#| label: fig-3
data <- read.csv(here::here("inputs/data/fig_5.csv"))

ggplot() + 
    geom_line(data = data, aes(x = year, y = brate_lesshs, color = "Less than High School")) +
    geom_line(data = data, aes(x = year, y = brate_hsgrad, color = "High School Graduate")) +
    geom_line(data = data, aes(x = year, y = brate_somecoll, color = "Some College")) +
    geom_line(data = data, aes(x = year, y = brate_colgrad, color = "College Graduate")) +
    theme_minimal() +
    labs(title = "US Birth Rates by Education Level (1992-2019)",
         x = "Year",
         y = "Birth Rate",
         color = "Education Level") +
    theme(legend.position = "bottom",
        legend.title = element_text(size=8),
        legend.text = element_text(size=7),
        legend.key.width = unit(0.5, "cm"), 
        legend.key.height = unit(.3, 'cm')) +
      guides(color = guide_legend(nrow = 2))

```
# 3.3 Marital Status vs Birth Rate

@@

```{r, fig.cap= "Birth rate of married vs unmarried women, rates trending towards each other over time"}
#| label: fig-4
data <- read.csv(here::here("inputs/data/fig_2e.csv"))

ggplot(data, aes(x = year)) + 
    geom_line(aes(y = brate_married, color = "Married")) +
    geom_line(aes(y = brate_unmarried, color = "Unmarried")) +
    geom_smooth(aes(y = brate_married), method = "lm", color = "red", linetype = "dashed", size = 0.3, se = FALSE) +
    geom_smooth(aes(y = brate_unmarried), method = "lm", color = "blue", linetype = "dashed", size = 0.3, se = FALSE) +
    theme_minimal() +
    labs(title = "US Birth Rates by Marital Status (1980-2020)",
         x = "Year",
         y = "Birth Rate",
         color = "Marital Status") +
    theme(legend.position = "bottom") 
```
# 4 Discussion: {#sec-discussion}

@@

\newpage
